Line of Sight Lensing Mass Reconstructions in the Millennium Simulation

### 2016-09-07 doc/thesis_swe.tex

While the effects of gravitational lensing can be dramatic, it is extremely rare for background galaxies to be perfectly aligned with large foreground mass to produce strongly lensed images. However, as Tyson et al. postulated in \cite{tyson_postulate} and confirmed in \cite{tyson_confirm}, background sources are also `weakly' lensed by lesser foreground mass nearby the line of sight (LOS) by what is now called galaxy-galaxy lensing. As the shape of most galaxies is only distorted by a few percent, and the instrinsic shape is not known, weak lensing must be detected statistically using a large number of sources.

Predicting accurate dark matter mass maps from weak lensing requires understanding the correlation between galaxies and dark matter structure. Numerous N-body simulations have predicted dark matter structure to have evolved from overdense regions in the early universe into `halos' of virialized matter that eventually collapsed from self-gravity \cite{halo_formation}, where galaxies could then form in the large gravitational potentials of the halos. However, there is still extensive reserach being done on the correct density profile of the halos (isothermal spheres \cite{isothermal_spheres}, NFW \cite{nfw}, BMO \cite{nfw_bmo}, Einasto \cite{einasto}), identifying the stellar mass - halo mass relation \cite{smhr}, abundance matching \cite{abundance_matching}, halo substructure \cite{halo_substructure}, and the lensing by LOS halos \cite{collett} and by dark matter structures not associated with galaxies \cite{mccully}. If a model for the distribution of dark matter in a region of foreground mass can accurately predict the statistical signal of the weak lensing of background sources, then the model can be used on galaxies in existing sky survey data to extrapolate the amount of dark matter in the region and construct large-scale maps of the dark matter in the universe.

@@

\begin{itemize}
\item Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct LOS dark matter mass structures using a halo model?
\item How many halos are needed to make these predictions?
\item How feasible is an inference of the Millennium Simulation's cosmological hyper-parameters, such as the stellar mass - halo mass relation, using all available halo data? 
\item How long does such an inference take?
\end{itemize}

@@

Before describing the developed model for LOS dark matter mass map reconstructions from weak lensing measurements, I briefely review a few topics in astronomy and cosmology that are relevant for this thesis such as the dark matter halo model, gravitational lensing, and correlation functions.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% FIGURE: Millennium Simulation Distribution plots (cosmic web, galaxy clusters with/without dark matter)
\begin{figure*}
    \centering
    \begin{subfigure}[H]{0.375\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ms_galaxies.jpg}
        %\captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{Galaxy distribution.}
        \label{ms_galaxies}
    \end{subfigure}
    ~
    %% Einstein Ring Figure
    \begin{subfigure}[H]{0.375\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ms_halos.jpg}
        %\captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{Halo distribution.}
        \label{ms_halos}
    \end{subfigure}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{A comparison of the galaxy and dark matter distribution for a single cluster in the Millennium Simulation. The galaxies tend to be located in the centers of clumps of dark matter called `halos', which suggests that galaxies at least partially trace out the underlying dark matter distribution. Taken from \cite{ms_figures}.}
    \label{ms_compare}
\end{figure*}

@@

The simplest way to model the relationship between galaxies and dark matter is by enveloping each galaxy in a spherically symmetric dark matter `halo' of a certain mass $M_h$. These halos extend far beyond the edge of the visible galaxy that they enclose, with the Milky Way's own halo estimated to have a radius of between 300,000 and 400,000 light-yeras \cite{milky_way_halo}. While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated locally by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

\noindent where the constant $\rho_0$ and the scale radius $R_s$ are parameters that vary from halo to halo \cite{nfw}. However, the NFW profile is not entierly physical as the total mass diverges as the density is integrated out to an infinite radius. Instead, this work uses a truncated NFW profile called the Baltz-Marshall-Oguri (BMO) profile given by
\begin{equation}\label{bmo_profile}
\rho_{BMO}(r)=\rho_{NFW}\cdot\left(\frac{r_t^2}{r^2+r_t^2}\right)^2
\end{equation}

@@

The halo model only accounts for dark matter structures associated with galaxies. However, this neglects other mass structures such as the filaments and voids that can be seen in Figure \ref{fig:ms_structure}. Modelling these features accurately would be difficult; we could try, for example, to model filaments by attaching cylinders of dark matter between massive clusters of galaxies. While an intriguing idea, rigorously testing for the appropriate filament density distributions, selection criteria, and frequency to statistically match simulations could constitute its own thesis.

@@

To solve for the desired smooth-component correction, we first need to calcualte ${\rho_m(z)}$ and ${\rho_h(z)}$ and then subtract:
\begin{equation}\label{smooth_correction2}
\rho_s(z)=\rho_m(z)-\rho_h(z).
\end{equation}

@@

While dark matter does not absorb or emit light, it does interact with light indirectly through gravitational lensing. The presence of matter curves spacetime locally which causes light to follow curved paths called geodesics. The deviation in path is usually negligable, but can be significant in regions containing clusters of galaxies and their massive dark matter halos. By observing how the magnitudes and shapes of distant galaxies are distorted when travelling through regions of matter, we can infer the masses of the dark matter halos and reconstruct the hidden dark matter mass distribution. %A simplified cartoon of this process is shown in Figure \ref{fig:lens_diagram}.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Relevance Distribution of all Halos
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figs-swe/thesis/relevance_distribution.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The mean number of relevant halos in a single realization of 1440 source galaxies, each with ${4836\pm104}$ halos per lightcone, determined by Equation \eqref{relevant_thresh} with a threshold mass and distance of ${M_T=10^{12}M_\odot}$ and ${R_T=10}$ kpc respectively. The errorbars correspond to the standard deviation of relevant halos accross all lightcones. As expected, most halos are either too far away from the LOS or not massive enough to make a significant contribution, or both. The number of halos falls logarithmically with the relevance limit and reaches a mean of one halo at a relevance of ${10^{-2}M_\odot/\text{Mpc}^3}$.}
    \label{fig:rel_dist}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

The volume and projected surface mass densities of each redshift slice for the same source realization as Figure \ref{fig:rel_dist} are shown in Figures \ref{fig:volume_densities} and \ref{fig:surface_densities} respectively. In each plot, the halo density at each redshift slice is plotted in green, the smooth-component density in blue, and universe mean density in black, as well as a histogram of the galaxies per bin in the background. In most redshift bins, the halo density is one to two orders of magnitude lower than than the universe mean density leading to a smooth-component density very close to that of the mean density. However, all mass densities are far below the critial density, plotted in red, and so well within the weak-lensing regime. The convergence distribution of each halo and the 100 smooth-components is plotted in Figure \ref{fig:kappa_dist}.
% This seems to be in opposition to Pace et al. in \cite{pace}, which found that halos constituted ??\% of simulation mass in ??.
%%------------------------------------------------------------------------------------------------------------------------------------------
%% Surface and Volume densities for smooth-component correction redshift slices
\begin{figure*}
    \centering
    % Volume
    \begin{subfigure}[H]{0.48\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/vol_densities.png}
        % \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{Smooth-component volume densities.}
        \label{fig:volume_densities}
    \end{subfigure}
    ~
    % Surface
    \begin{subfigure}[H]{0.48\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/surf_densities.png}
        % \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{Smooth-component projected surface densities.}
        \label{fig:surface_densities}
    \end{subfigure}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The surface and volume densities for each of the 100 smooth-component correction and halo densities in a lightcone with radius $R_L=6$ arcminutes. The green line is the halo density of each slice and the blue line is the calculated smooth-component density such that the two densities sum to the universe mean density plotted in black. The number of halos in each redshift bin is plotted as a histogram in the background. The red line in plot (b) is the ciritical surface density.}
    \label{densities}
\end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Pangloss Lightcone Data Plots
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{figs-swe/thesis/lightcone_plots.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{Four different views of a lightcone centered on the LOS of a particular background source.  The top left figure plots the positions of the halos with the size proportional to their galaxy brightness in the $i$-band, the center figure plots the angular size of foreground halos with red and blue regions representing the NFW scale radius and virial radius of each halo respectively, the top right figure plots the $\kappa$ contributions of each halo, and the bottom plots the positions of each halo along the redshift axis with size proportional to $\kappa$ contribution.}
    \label{fig:lightcone}
\end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

Figure \ref{fig:lightcone} displays different views an example lightcone where the foreground halos are plotted as circles whose size are proportional to their galaxy brightness (top left), angular size (top center), \texttt{Pangloss} predicted convergence contributions (top right), and halo mass (bottom). Comparing the center and top right figures demonstrates the importance of proximity to the LOS for significant $\kappa$ contributions. Figure \ref{fig:field2} shows a foreground cluster in a 1.44 arcmin$^2$ section of the Millennium Simulation along with the ray-traced convergence (plotted in inverse greyscale) and shear maps (plotted as red shear sticks) from \cite{ray_tracing}. Figure \ref{fig:field3} displays the generated background ellipticities in blue, the `true' lensed ellipticities in green, and the halo model lensed ellipticities in purple. While there is a visible difference between the lensed ellipticities, both align tangentially around the foreground mass as expected.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Lensing of a few background sources
\begin{figure*}
    \centering
    % Convergence / shear map, background source ellipticities
    \begin{subfigure}[H]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/field2.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The ray-traced convergence and shear in a 1.44 arcmin$^2$ field of the Millennium Simulation, as calculated from \cite{ray_tracing}. The convergence, plotted in inverse greyscale, is proportional to the projected mass density and representes foreground mass. The shear field is plotted as sparsely-sampled red sticks that have size proportional to their magnitude. A few of the generated background sources are plotted as blue ellipticitiy sticks with length proportional to their eccentricities.}
        \label{fig:field2}
    \end{subfigure}
     ~
    % Lensed ellipticities
    \begin{subfigure}[H]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/field3.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The same 1.44 arcmin$^2$ field as \ref{fig:field2}, but now with the `true' lensed ellipticities in green, the halo model predicted lensed ellipticities in purple, and a few of the most massive foreground halos plotted as orange circles. Lensed ellipticity predictions tend to be more accurate further from the lensing mass, with significant differences near the centers of massive objects. All are sheared tangentially around the massive structures as expected.}
        \label{fig:field3}
    \end{subfigure}
    \caption{}
    \label{field}
\end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
% Ellipticity-Ellipticity Correlation
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/gg_corr.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The $\varepsilon$-$\varepsilon$ correlation function for 1440 sources at separation distances from 0.1 to 2 arcminutes using lightcones with radii of $R_L=6$ arcminutes, where green is the `true' ray-traced correlation and purple is the halo model predicted correlation. While the cross-component $\xi_\times(\Delta\theta)$ for both is consistent with zero accross all scales as expected, the halo model systematically underpredicts the correlation on scales larger than 0.2 arcminutes.}
        \label{fig:gg_corr}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/gg_progression.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The $\varepsilon$-$\varepsilon$ correlation functions of 1440 sources for a series of halo model frameworks with increasing lightcone radii from $R_L=1$ to $R_L=8$ arcminutes, all plotted in purple with various line styles. These are compared to the same ray-traced correlation from Figure \ref{fig:gg_corr}. The predicted correlation seems to converge to the ray-traced value with larger $R_L$ as more halo contributions are accounted for.}
        \label{fig:gg_corr_series}
    \end{subfigure}
    \caption{}
\end{figure}
%------------------------------------------------------------------------------------------------------------------------------------------

The first test of the halo model was with the $\varepsilon$-$\varepsilon$ correlation function with a lightcone radius of 6 arcminutes and all $4836\pm104$ halos per lightcone which is given in Figure \ref{fig:gg_corr}. The statistic measured the average correlation between pairs of ellipticities at separation distances between 0.1 and 2 arcminutes. The cross-component of the correlation function $\xi_\times$ shows no significant deviation from zero as expected, as gravitational lensing only shears galaxies tangentially. For the $\xi_+$ component, the ray-traced (plotted in green) and halo model (plotted in purple) are in relative agreement from seperation scales of 0.1 to 0.2 arcminutes, but there is a systematic underprediction of correlation by the halo model on scales larger than 0.2 arcminutes by an average 31\% percent error. This result is consistent with Pace et al. in \cite{pace} which found that a purely halo model underpredicted the lensing correlation on average by 40\%. This may be indicative that not enough halo mass is being used in the lensing prediction, the halo model does not adequately address large-scale mass structures, or that there is significant dark matter mass not correlated with galaxies.  These issues are discussed further in Section \ref{issues}.

@@

The galaxy-mass correlation function was similarly re-calculated at increasing lightcone radii from $R=1$ to $R=8$ arcminutes with the result in Figure \ref{fig:ng_corr_series}. As with the $\varepsilon$-$\varepsilon$ correlation, structure larger than $\sim$1 arcminute is captured much better by increasingly large lightcone radii as expected. However, there is little to no improvement of the galaxy-mass correlation on scales smaller than 0.3 arcminutes which futher points to the need to model strong lensing.

@@

While correlation functions can describe how well the lensing predictions are being made globally, it is still instructive to analyze the difference between the ray-traced and halo model projected mass maps. This was done by selecting a region of foreground mass, calculating the ray-traced and \texttt{Pangloss}-predicted convergence of a uniform grid of background sources in the field, and binning the convergence maps to a desired resolution. Convergence is proportional to projected surface mass density, so comparing the two convergence maps is equivalent to comparing their respecitve projected mass distributions.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
% Mean lightcone CPU time
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/profile_bar.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The mean CPU runtime per lightcone for various runs with different (and cumulative) code optimizations described in Section \ref{comp_perform}. The cumulative changes account for a 59\% decrease in the mean CPU time and a halving of the initial variation. On average, the lensing prediction for a single lightcone can now be made in $57\pm8$ ms.}
    \label{fig:cpu_plot}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

The distribution of total CPU time in significant parts of the lensing calculation for run 6 are as follows: 24.0\% of total time calculating individual halo convergence contributions where 14.4\% is spent in lensing table lookup functions, 65.1\% of CPU time was spent combining all convergence contributions where 62.1\% is spent on the smooth-component correction, and the majority of the remaining time spent on writing the output \texttt{Astropy} columns. However, time spent accross all components of the lensing calculation was dominated by \texttt{Astropy} column use with nearly 55\% of total time spent initializing, copying, or summing column elements.

@@

Perhaps the most common method of parameter estimation of statistical models is called maximum-likelihood estimation (MLE). The likelihood is a function of the parameters of a statistical model that measures the probability of a data set or outcomes given the model parameters \cite{bayes_in_sky}. It is the mathematical answer to the question, ``\textit{Given a particular set of parameters, what is the probability that this data set could have occured?}" The form of the likelihood depends on the properties and circumstances of the data, but generally has the form of the product of the probability density function $f$ of each individual observation $x_i$ given model parameters $\vec{\theta}$:
\begin{equation}
\mathcal{L}(\vec{\theta},x_1,x_2,\ldots,x_n)=\prod_{i=1}^nf(x_i|\vec{\theta}).
\end{equation}

@@

However, exploring the parameter space efficiently is a significant problem in MLE, especially in a dataset such as this which there are potentially millions or even billions of parameters (i.e. halo masses). For this thesis, I only compute the log-likelihood for a small handful of parameter sets to gague how feasible such an inference would be in future work with far more computational power.

>>>>

While the effects of gravitational lensing can be dramatic, it is extremely rare for background galaxies to be perfectly aligned with large foreground mass to produce strongly lensed images. However, as Tyson et al. postulated in \cite{tyson_postulate} and confirmed in \cite{tyson_confirm}, background sources are also `weakly' lensed by lesser foreground mass nearby the line of sight (LOS) by what is now called galaxy-galaxy lensing. As the shape of most galaxies is only distorted by a few percent, and the intrinsic shape is not known, weak lensing must be detected statistically using a large number of sources.

Predicting accurate dark matter mass maps from weak lensing requires understanding the correlation between galaxies and dark matter structure. Numerous N-body simulations have predicted dark matter structure to have evolved from overdense regions in the early universe into `halos' of virialized matter that eventually collapsed from self-gravity \cite{halo_formation}, where galaxies could then form in the large gravitational potentials of the halos. However, there is still extensive research being done on the correct density profile of the halos (isothermal spheres \cite{isothermal_spheres}, NFW \cite{nfw}, BMO \cite{nfw_bmo}, Einasto \cite{einasto}), identifying the stellar mass - halo mass relation \cite{smhr}, abundance matching \cite{abundance_matching}, halo substructure \cite{halo_substructure}, and the lensing by LOS halos \cite{collett} and by dark matter structures not associated with galaxies \cite{mccully}. If a model for the distribution of dark matter in a region of foreground mass can accurately predict the statistical signal of the weak lensing of background sources, then the model can be used on galaxies in existing sky survey data to extrapolate the amount of dark matter in the region and construct large-scale maps of the dark matter in the universe.

@@

\begin{itemize}
\item Under the best possible circumstances of having noiseless redshifts and halo masses, how accurately can I reconstruct LOS dark matter mass structures using a halo model?
\item How many halos are needed to make these predictions?
\item How feasible is an inference of the Millennium Simulation's cosmological hyperparameters, such as the stellar mass - halo mass relation, using all available halo data? 
\item How long does such an inference take?
\end{itemize}

@@

Before describing the developed model for LOS dark matter mass map reconstructions from weak lensing measurements, I briefly review a few topics in astronomy and cosmology that are relevant for this thesis such as the dark matter halo model, gravitational lensing, and correlation functions.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% FIGURE: Millennium Simulation Distribution plots (cosmic web, galaxy clusters with/without dark matter)
\begin{figure*}
    \centering
    \begin{subfigure}[H]{0.375\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ms_galaxies.jpg}
        %\captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{Galaxy distribution.}
        \label{ms_galaxies}
    \end{subfigure}
    ~
    %% Einstein Ring Figure
    \begin{subfigure}[H]{0.375\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/ms_halos.jpg}
        %\captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{Halo distribution.}
        \label{ms_halos}
    \end{subfigure}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{A comparison of the galaxy and dark matter distribution for a single cluster in the Millennium Simulation. Galaxies tend to be located in the centers of clumps of dark matter called `halos', which suggests that galaxies at least partially trace out the underlying dark matter distribution. Taken from \cite{ms_figures}.}
    \label{ms_compare}
\end{figure*}

@@

The simplest way to model the relationship between galaxies and dark matter is by enveloping each galaxy in a spherically symmetric dark matter `halo' of a certain mass $M_h$. These halos extend far beyond the edge of the visible galaxy that they enclose, with the Milky Way's own halo estimated to have a radius of between 300,000 and 400,000 light-years \cite{milky_way_halo}. While the density profile of the halos may be complex, numerous simulations have shown that it can be well approximated locally by the Navarro-Frenk-White (NFW) profile which has the form
\begin{equation}\label{nfw_profile}
\rho_{NFW}(r)=\frac{\rho_0}{\frac{r}{R_s}\left(1+\frac{r}{R_s}\right)^2}
\end{equation}

\noindent where the constant $\rho_0$ and the scale radius $R_s$ are parameters that vary from halo to halo \cite{nfw}. However, the NFW profile is not entirely physical as the total mass diverges as the density is integrated out to an infinite radius. Instead, this work uses a truncated NFW profile called the Baltz-Marshall-Oguri (BMO) profile given by
\begin{equation}\label{bmo_profile}
\rho_{BMO}(r)=\rho_{NFW}\cdot\left(\frac{r_t^2}{r^2+r_t^2}\right)^2
\end{equation}

@@

The halo model only accounts for dark matter structures associated with galaxies. However, this neglects other mass structures such as the filaments and voids that can be seen in Figure \ref{fig:ms_structure}. Modeling these features accurately would be difficult; we could try, for example, to model filaments by attaching cylinders of dark matter between massive clusters of galaxies. While an intriguing idea, rigorously testing for the appropriate filament density distributions, selection criteria, and frequency to statistically match simulations could constitute its own thesis.

@@

To solve for the desired smooth-component correction, we first need to calculate ${\rho_m(z)}$ and ${\rho_h(z)}$ and then subtract:
\begin{equation}\label{smooth_correction2}
\rho_s(z)=\rho_m(z)-\rho_h(z).
\end{equation}

@@

While dark matter does not absorb or emit light, it does interact with light indirectly through gravitational lensing. The presence of matter curves spacetime locally which causes light to follow curved paths called geodesics. The deviation in path is usually negligible, but can be significant in regions containing clusters of galaxies and their massive dark matter halos. By observing how the magnitudes and shapes of distant galaxies are distorted when traveling through regions of matter, we can infer the masses of the dark matter halos and reconstruct the hidden dark matter mass distribution. %A simplified cartoon of this process is shown in Figure \ref{fig:lens_diagram}.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Relevance Distribution of all Halos
\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{figs-swe/thesis/relevance_distribution.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The mean number of relevant halos in a single realization of 1440 source galaxies, each with ${4836\pm104}$ halos per lightcone, determined by Equation \eqref{relevant_thresh} with a threshold mass and distance of ${M_T=10^{12}M_\odot}$ and ${R_T=10}$ kpc respectively. The error bars correspond to the standard deviation of relevant halos across all lightcones. As expected, most halos are either too far away from the LOS or not massive enough to make a significant contribution, or both. The number of halos falls logarithmically with the relevance limit and reaches a mean of one halo at a relevance of ${10^{-2}M_\odot/\text{Mpc}^3}$.}
    \label{fig:rel_dist}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

The volume and projected surface mass densities of each redshift slice for the same source realization as Figure \ref{fig:rel_dist} are shown in Figures \ref{fig:volume_densities} and \ref{fig:surface_densities} respectively. In each plot, the halo density at each redshift slice is plotted in green, the smooth-component density in blue, and universe mean density in black, as well as a histogram of the galaxies per bin in the background. In most redshift bins, the halo density is one to two orders of magnitude lower than than the universe mean density leading to a smooth-component density very close to that of the mean density. However, all mass densities are far below the critical density, plotted in red, and so well within the weak-lensing regime. The convergence distribution of each halo and the 100 smooth-components is plotted in Figure \ref{fig:kappa_dist}.
% This seems to be in opposition to Pace et al. in \cite{pace}, which found that halos constituted ??\% of simulation mass in ??.
%%------------------------------------------------------------------------------------------------------------------------------------------
%% Surface and Volume densities for smooth-component correction redshift slices
\begin{figure*}
    \centering
    % Volume
    \begin{subfigure}[H]{0.48\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/vol_densities.png}
        % \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{Smooth-component volume densities.}
        \label{fig:volume_densities}
    \end{subfigure}
    ~
    % Surface
    \begin{subfigure}[H]{0.48\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/surf_densities.png}
        % \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{Smooth-component projected surface densities.}
        \label{fig:surface_densities}
    \end{subfigure}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The surface and volume densities for each of the 100 smooth-component correction and halo densities in a lightcone with radius $R_L=6$ arcminutes. The green line is the halo density of each slice and the blue line is the calculated smooth-component density such that the two densities sum to the universe mean density plotted in black. The number of halos in each redshift bin is plotted as a histogram in the background. The red line in plot (b) is the critical surface density.}
    \label{densities}
\end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Pangloss Lightcone Data Plots
\begin{figure*}
    \centering
    \includegraphics[width=0.8\textwidth]{figs-swe/thesis/lightcone_plots.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{Four different views of a lightcone with radius $R_L=6$ arcminutes centered on the LOS of a particular background source containing 4829 halos.  The top left figure plots the positions of the halos with the size proportional to their galaxy brightness in the $i$-band, the center figure plots the angular size of foreground halos with red and blue regions representing the NFW scale radius and virial radius of each halo respectively, the top right figure plots the $\kappa$ contributions of each halo, and the bottom plots the positions of each halo along the redshift axis with size proportional to $\kappa$ contribution.}
    \label{fig:lightcone}
\end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

Figure \ref{fig:lightcone} displays different views an example lightcone where the foreground halos are plotted as circles whose size are proportional to their galaxy brightness (top left), angular size (top center), \texttt{Pangloss} predicted convergence contributions (top right), and halo mass (bottom). Comparing the center and top right figures demonstrates the importance of proximity to the LOS for significant $\kappa$ contributions. Figure \ref{fig:field2} shows a foreground cluster in a 1.44 arcmin$^2$ section of the Millennium Simulation along with the ray-traced convergence (plotted in inverse grayscale) and shear maps (plotted as red shear sticks) from \cite{ray_tracing}. Figure \ref{fig:field3} displays the generated background ellipticities in blue, the `true' lensed ellipticities in green, and the halo model lensed ellipticities in purple. While there is a visible difference between the lensed ellipticities, both align tangentially around the foreground mass as expected.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
%% Lensing of a few background sources
\begin{figure*}
    \centering
    % Convergence / shear map, background source ellipticities
    \begin{subfigure}[H]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/field2.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The ray-traced convergence and shear in a 1.44 arcmin$^2$ field of the Millennium Simulation, as calculated from \cite{ray_tracing}. The convergence, plotted in inverse grayscale, is proportional to the projected mass density and represents foreground mass. The shear field is plotted as sparsely-sampled red sticks that have size proportional to their magnitude. A few of the generated background sources are plotted as blue ellipticity sticks with length proportional to their eccentricities.}
        \label{fig:field2}
    \end{subfigure}
     ~
    % Lensed ellipticities
    \begin{subfigure}[H]{0.45\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/field3.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The same 1.44 arcmin$^2$ field as \ref{fig:field2}, but now with the `true' lensed ellipticities in green, the halo model predicted lensed ellipticities in purple, and a few of the most massive foreground halos plotted as orange circles. Lensed ellipticity predictions tend to be more accurate further from the lensing mass, with significant differences near the centers of massive objects. All are sheared tangentially around the massive structures as expected.}
        \label{fig:field3}
    \end{subfigure}
    \caption{}
    \label{field}
\end{figure*}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
% Ellipticity-Ellipticity Correlation
\begin{figure}
    \centering
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/gg_corr.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The $\varepsilon$-$\varepsilon$ correlation function for 1440 sources at separation distances from 0.1 to 2 arcminutes using lightcones with radii of $R_L=6$ arcminutes, where green is the `true' ray-traced correlation and purple is the halo model predicted correlation. While the cross-component $\xi_\times(\Delta\theta)$ for both is consistent with zero across all scales as expected, the halo model systematically underpredicts the correlation on scales larger than 0.2 arcminutes.}
        \label{fig:gg_corr}
    \end{subfigure}
    ~~
    \begin{subfigure}{0.475\textwidth}
        \includegraphics[width=\textwidth]{figs-swe/thesis/gg_progression.png}
        \captionsetup{justification=raggedright,singlelinecheck=false}
        \caption{The $\varepsilon$-$\varepsilon$ correlation functions of 1440 sources for a series of halo model frameworks with increasing lightcone radii from $R_L=1$ to $R_L=8$ arcminutes, all plotted in purple with various line styles. These are compared to the same ray-traced correlation from Figure \ref{fig:gg_corr}. The predicted correlation seems to converge to the ray-traced value with larger $R_L$ as more halo contributions are accounted for.}
        \label{fig:gg_corr_series}
    \end{subfigure}
    \caption{}
\end{figure}
%------------------------------------------------------------------------------------------------------------------------------------------

The first test of the halo model was with the $\varepsilon$-$\varepsilon$ correlation function with a lightcone radius of 6 arcminutes and all $4836\pm104$ halos per lightcone which is given in Figure \ref{fig:gg_corr}. The statistic measured the average correlation between pairs of ellipticities at separation distances between 0.1 and 2 arcminutes. The cross-component of the correlation function $\xi_\times$ shows no significant deviation from zero as expected, as gravitational lensing only shears galaxies tangentially. For the $\xi_+$ component, the ray-traced (plotted in green) and halo model (plotted in purple) are in relative agreement from separation scales of 0.1 to 0.2 arcminutes, but there is a systematic underprediction of correlation by the halo model on scales larger than 0.2 arcminutes by an average 31\% error. This result is consistent with Pace et al. in \cite{pace} which found that a purely halo model underpredicted the lensing correlation on average by 40\%. This may be indicative that not enough halo mass is being used in the lensing prediction, the halo model does not adequately address large-scale mass structures, or that there is significant dark matter mass not correlated with galaxies.  These issues are discussed further in Section \ref{issues}.

@@

The galaxy-mass correlation function was similarly re-calculated at increasing lightcone radii from $R=1$ to $R=8$ arcminutes with the result in Figure \ref{fig:ng_corr_series}. As with the $\varepsilon$-$\varepsilon$ correlation, structure larger than $\sim$1 arcminute is captured much better by increasingly large lightcone radii as expected. However, there is little to no improvement of the galaxy-mass correlation on scales smaller than 0.3 arcminutes which further points to the need to model strong lensing.

@@

While correlation functions can describe how well the lensing predictions are being made globally, it is still instructive to analyze the difference between the ray-traced and halo model projected mass maps. This was done by selecting a region of foreground mass, calculating the ray-traced and \texttt{Pangloss}-predicted convergence of a uniform grid of background sources in the field, and binning the convergence maps to a desired resolution. Convergence is proportional to projected surface mass density, so comparing the two convergence maps is equivalent to comparing their respective projected mass distributions.

@@

%%------------------------------------------------------------------------------------------------------------------------------------------
% Mean lightcone CPU time
\begin{figure}
    \centering
    \includegraphics[width=0.475\textwidth]{figs-swe/thesis/profile_bar.png}
    \captionsetup{justification=raggedright,singlelinecheck=false}
    \caption{The mean CPU runtime per lightcone with radius $R_L=6$ arcminutes for various runs with different (and cumulative) code optimizations described in Section \ref{comp_perform}. The cumulative changes account for a 59\% decrease in the mean CPU time and a halving of the initial variation. On average, the lensing prediction for a single lightcone can now be made in $57\pm8$ ms.}
    \label{fig:cpu_plot}
\end{figure}
%%------------------------------------------------------------------------------------------------------------------------------------------

@@

The distribution of total CPU time in significant parts of the lensing calculation for run 6 are as follows: 24.0\% of total time calculating individual halo convergence contributions where 14.4\% is spent in lensing table lookup functions, 65.1\% of CPU time was spent combining all convergence contributions where 62.1\% is spent on the smooth-component correction, and the majority of the remaining time spent on writing the output \texttt{Astropy} columns. However, time spent across all components of the lensing calculation was dominated by \texttt{Astropy} column use with nearly 55\% of total time spent initializing, copying, or summing column elements.

@@

Perhaps the most common method of parameter estimation of statistical models is called maximum-likelihood estimation (MLE). The likelihood is a function of the parameters of a statistical model that measures the probability of a data set or outcomes given the model parameters \cite{bayes_in_sky}. It is the mathematical answer to the question, ``\textit{Given a particular set of parameters, what is the probability that this data set could have occurred?}" The form of the likelihood depends on the properties and circumstances of the data, but generally has the form of the product of the probability density function $f$ of each individual observation $x_i$ given model parameters $\vec{\theta}$:
\begin{equation}
\mathcal{L}(\vec{\theta},x_1,x_2,\ldots,x_n)=\prod_{i=1}^nf(x_i|\vec{\theta}).
\end{equation}

@@

However, exploring the parameter space efficiently is a significant problem in MLE, especially in a dataset such as this which there are potentially millions or even billions of parameters (i.e. halo masses). For this thesis, I only compute the log-likelihood for a small handful of parameter sets to gauge how feasible such an inference would be in future work with far more computational power.
